Fast and scalable Gaussian process modeling with applications to astronomical time series \vspace{-3\baselineskip}

### 2016-12-30 paper/ms.tex
typo
another typo
acks
Small corrections; p -> J; p_0 -> J_0

It turns out that this kernel function can be made even more general by
introducing complex parameters $a_j \to a_j+i\,b_j$ and $c_j \to c_j+i\,d_j$.
In this case, the covariance function becomes
\begin{eqnarray}\eqlabel{genrp-kernel-complex}
k_\bvec{\alpha}(\tau_{nm}) = \sigma_n^2\,\delta_{nm} +
    \sum_{j=1}^N &&\left[
    \frac{1}{2}(a_j + i\,b_j)\,\exp\left(-(c_j+i\,d_j)\,\tau_{nm}\right)
        \right. \nonumber\\
    &&+\left.
    \frac{1}{2}(a_j - i\,b_j)\,\exp\left(-(c_j-i\,d_j)\,\tau_{nm}\right)
\right]
\end{eqnarray}
and, for this function, \eq{gp-likelihood} can still be solved with
$\mathcal{O}(J^2\,N)$ operations.
The details of this method and a few implementation considerations are
discussed in \app{implementation}.

\eq{genrp-kernel-complex} can also be expressed as
\begin{eqnarray}\eqlabel{genrp-kernel}
k_\bvec{\alpha}(\tau_{nm}) = \sigma_n^2\,\delta_{nm} +
    \sum_{j=1}^J &&\left[
    a_j\,\exp\left(-c_j\,\tau_{nm}\right)\,\cos\left(d_j\,\tau_{nm}\right)
        \right.\nonumber\\
    &&+ \left.
    b_j\,\exp\left(-c_j\,\tau_{nm}\right)\,\sin\left(d_j\,\tau_{nm}\right)
\right] \quad.
\end{eqnarray}
The Fourier transform of this covariance function is the power spectral
density (PSD) of the model and it is given by
\begin{eqnarray}\eqlabel{genrp-psd}
S(\omega) = \sum_{j=1}^p \sqrt{\frac{2}{\pi}}
\frac{(a_j\,c_j+b_j\,d_j)\,({c_j}^2+{d_j}^2)+(a_j\,c_j-b_j\,d_j)\,\omega^2}
{\omega^4+2\,({c_j}^2-{d_j}^2)\,\omega^2+({c_j}^2+{d_j}^2)^2}\quad.
\end{eqnarray}
The physical interpretation of this model isn't immediately obvious and we
will return to a more general discussion of the physical intuition in a moment
but we can start with a discussion of some useful special cases of this model.

If we set the imaginary amplitude $b_j$ for some component $j$ to zero, that
term of \eq{genrp-kernel} becomes
\begin{eqnarray}
k_j(\tau_{nm}) =
    a_j\,\exp\left(-c_j\,\tau_{nm}\right)\,\cos\left(d_j\,\tau_{nm}\right)
\end{eqnarray}
and the PSD for the this component is
\begin{eqnarray}
S_j(\omega) = \frac{1}{\sqrt{2\,\pi}}\,\frac{a_j}{c_j}\,\left[
    \frac{1}{1+\left(\frac{\omega-d_j}{c_j}\right)^2} +
    \frac{1}{1+\left(\frac{\omega+d_j}{c_j}\right)^2}
\right] \quad.
\end{eqnarray}
This PSD is the sum of two Lorentzian or Cauchy distributions with width
$c_j$ centered on $\omega = \pm d_j$.
This model can be interpreted intuitively as a quasiperiodic oscillator with
amplitude $A_j = a_j$, quality factor $Q_j = c_j$, and period $P_j = 2\,\pi /
d_j$.

@@

In the case of $p$ damped sinusoids we can check for negative values of the PSD by solving for the
roots of the power spectrum, abbreviating with $z = \omega^2$:
\begin{equation}
P(\omega)=  \sum_{j=1}^{p} \frac{q_j z + r_j}{z^2+s_jz + t_j} = 0
\end{equation}
where
\begin{eqnarray}
q_j &=& a_jc_j-b_jd_j\\
r_j &=& (d_j^2+c_j^2)(b_jd_j+a_jc_j)\\
s_j &=& 2(c_j^2-d_j^2)\\
t_j &=& (c_j^2+d_j^2)^2.
\end{eqnarray}
The denominators of each term are positive, so we can multiply through by $\Pi_j \left(z^2+s_jz + t_j\right)$ yielding:
\begin{equation}
P_0(z) = \sum_{j=1}^{p} (q_j z + r_j)\Pi_{k \ne j}\left(z^2+s_kz + t_k\right) = 0,
\end{equation}
which is a polynomial with order $p_{ord}=2(p-1)+1$.  With $p=2$, this yields a cubic equation
which may be solved exactly for the roots.

@@

{\item Substituting $Q = 1/2$, \eq{sho-psd} becomes
\begin{eqnarray}
S(\omega) =
\sqrt{\frac{2}{\pi}}\,\frac{S_0}{\left[(\omega/\omega_0)^2+1\right]^2}
\end{eqnarray}
with the corresponding covariance function
\begin{eqnarray}
k(\tau) &=& \lim_{f \to 0}\,
    \frac{1}{2}\,S_0\,\omega_0\,e^{-\omega_0\,\tau}\,
    \left[\cos(f\,\tau) + \frac{\omega_0}{f}\,\sin(f\,\tau)\right] \\
&=& \frac{1}{2}\,S_0\,\omega_0\,e^{-\omega_0\,\tau}\,[1-\omega_0\,\tau] \quad.
\end{eqnarray}
This covariance function is also known as the Mat\'ern-3/2 function
\citep{Rasmussen:2006}.
}

@@

%The kernel is modeled as a sum of damped harmonics,
%\begin{equation}
%k_j(\tau) = a_0 \delta(\tau) + \sum_{j=1}^{N_{kernel}} a_j \exp{(-b_j \tau)} \cos{(-d_j \tau)} + b_j \exp{(-b_j \tau)} \sin{(-d_j \tau)},
%\end{equation}
%where $\tau > 0$ is a time-offset between data points, $c_j$ represents the decay
%in correlation with time and $d_j$ represents oscillations of period $P = 2\pi/d_j$, and
%$a_0 = \sigma^2$ is a white-noise component with $\delta(0) = 1$ and zero otherwise.  Formally
%the $\delta(\tau)$ function is a Kronecker delta function for which $\tau$ is the midpoint in
%time measured over a finite duration exposure time, and $\sigma$ is the standard deviation
%of the measured times if the correlated components are zero.

@@

One advantage of the Mat\'ern 3/2 kernel is that it is smooth across
$\tau=0$ since its derivative is zero at $\tau=0$.  The exponential kernel {\it does not}
have this property, which leads to qualitatively different noise properties on small
timescales due to the sharpness of the exponential kernel at $\tau=0$.  The Mat\'ern 3/2 kernel
turns out to be similar to the observed properties of stellar activity and granulation, and thus the approximate
formula is simultaneously physically relevant and computationally convenient.
The power spectrum of a single exponential scales as $f^{-2}$ at high frequency,
while the Mat\'ern kernel scales as $f^{-4}$ which more closely matches the power
spectrum due to activity and granulation in the Sun \citep{2004A&A...414.1139A,2009A&A...495..979M}.
\item {\bf Mat\'ern 5/2 kernel.}  The Mat\'ern 5/2 kernel is given by
\begin{equation}
k_{5/2}(t) = C_{5/2}(y) = \sigma^2 (1+y+y^2/3)e^{-y},
\end{equation}
where $y = \sqrt{5}\vert \tau/\tau_0\vert$.  The Fourier transform is:
\begin{equation}
\hat k_{5/2}(\omega) = \frac{2\tau_0 10^{5/2}}{3\pi^{1/2}} (5+\tau_0^2\omega^2)^{-3}.
\end{equation}
This kernel may be approximated well by the sum of four exponential kernels
\begin{eqnarray}
k_{5/2}(\tau) &\approx& -147.6965573 e^{-1.0097610 y} + 85.8875034 e^{-1.0621734 y}\\
 &-&4.6540017 e^{-0.8353745 y} + 67.4630744 e^{-0.9160394 y},
\end{eqnarray}
which is accurate to $<4\times 10^{-6}$ for $y \le 10$.
The Fourier transform has a steeper dependence at large frequencies, $f^{-6}$, and as we are not aware of a stellar
power spectrum that requires a steeper spectrum than this, we forego approximating further Mat\'ern kernels.
\item {\bf Exponential-sine-square kernel.}  This kernel is a periodic, positive valued kernel
which has been used to model quasi-periodic stellar variability due to star spots.   It is
given by $E(w,\Gamma) = e^{-\Gamma \sin^2{w}}$ where $w = \frac{\pi}{P}\vert \tau \vert$, which may be
expanded as a series in $\cos{(2kw)} = \cos{\frac{2\pi}{P} k \vert \tau \vert)}$,
with $k \in \mathbb{N}$.  We have expanded this to $k=4$, and fit for the coefficients, giving:
\begin{equation}
E(w,\Gamma) \approx \sum_{k=0}^4 \left(c_{k,1}+c_{k,2} e^{-c_{k,4}\Gamma^{c_{k,3}}}\right) \cos{(2kw)},
\end{equation}
with $c_{0,1}=0.24999, c_{0,2}=0.752803, c_{0,3}=0.960477, c_{0,4}=0.643393$ for $k=0$,
$c_{1,1}=0.450851, c_{1,2}=-0.445338, c_{1,3}=0.960477, c_{1,4}= 1.18102$ for $k=1$,
$c_{2,1}=0.206593, c_{2,2}=-0.207361, c_{2,3}=1.65902, c_{2,4}= 0.210641$ for $k=2$,
$c_{3,1}=0.0728449, c_{3,2}=-0.0729381, c_{3,3}=2.43149, c_{3,4}= 0.0471224$ for $k=3$, and
$c_{4,1}=0.019204, c_{4,2}=-0.0192129, c_{4,3}=3.27977, c_{4,4}= 0.0114198$ for $k=4$.
This approximation is accurate to better than 0.6\% for $\Gamma < 3$, with a standard
deviation of 0.1\%.
\end{itemize}

@@

%\subsection{Derivative of likelihood with respect to Kernel parameters}
%
%The derivative of the likelihood can be computed in $O(N)$ operations as follows.
%
%The derivative of the log likelihood with respect to model parameters ${\bf x}$ is given by:
%\begin{equation}
%    \frac{\partial \ln{\cal L}}{\partial x_i} = (\ln{\cal L})^\prime = -\frac{1}{2}\sum_i \frac{K_{ii}^\prime}{K_{ii}}
%    - \frac{1}{2} {\bf y}^T {\bf K}^{-1} {\bf K}^\prime {\bf K}^{-1} {\bf y},
%\end{equation}
%where $^\prime$ means the derivative wrt $x_i$.
%The kernel may be written as:
%\begin{equation}
%    K_{ij}({\bf \alpha},{\bf \beta}) = \left\{
%    \begin{array}{cc}
%       w_i + \sum_{l=1}^p \alpha_l & j = i \\
%       \sum_{l=1}^p \alpha_l \exp{(-\beta_l |t_i-t_j|)}  & j \ne i
%    \end{array}
%    \right.
%\end{equation}
%Since this equation is the sum of $p$ semi-separable matrices, then applying the product rule to compute the derivative will give a single semi-separable matrix for the derivative with respect to each kernel parameter.
%
%Now, letting ${\bf x} = (w,{\bf \alpha},{\bf \beta})$, then the derivative may be written as:
%\begin{eqnarray}
%    \frac{\partial K_{ii}}{\partial w} &=& \delta_{ij}\\
%    \frac{\partial K_{ij}}{\partial \alpha_k} &=&
%     \left\{
%    \begin{array}{cc}
%       1 & j = i \\
%        \exp{(-\beta_k |t_i-t_j|)}  & j \ne i
%    \end{array}
%    \right.\\
%    \frac{\partial K_{ij}}{\partial \beta_k} &=&
%     \left\{
%    \begin{array}{cc}
%       0  & j = i \\
%        -\alpha_k|t_i-t_j|\exp{(-\beta_k |t_i-t_j|)}  & j \ne i
%    \end{array}
%    \right.
%\end{eqnarray}
%The first term in the derivative of the likelihood can be computed in $O(N)$
%evaluations, while the second term can be computed in three steps:
%\begin{itemize}
%\item First, solve ${\bf z_1} = {\bf K}^{-1}{\bf y}$ with the GRP solver.
%\item Next, create an e
%\end{itemize}
%Writing this term out:

@@

%\section{Prediction/interpolation}
%
%Precition/interpolation involves constructing a matrix of covariances
%between the (noisy) training set and the times at which we wish to
%predict or interpolate data points.   If our training times are given by
%${\bf t}$, while the predicted times are given by ${\bf t}_*$, then
%mean values at the predicted times are given by (RW 2.23):
%\begin{eqnarray}
%\bar f_{*,k} &=& \sum_j k(t_{*,k},t_j) b_j,\\
%&=& \sum_j \sum_p \alpha_p e^{-\beta_p \vert t_{*,k}-t_j\vert} b_j,\\
%&=& \sum_p \alpha_p \sum_j e^{-\beta_p \vert t_{*,k}-t_j\vert} b_j,\\
%y_k &=& \sum_j k(t_j,t_k) h_j,
%\end{eqnarray}
%where the latter equation may be solved for ${\bf h}$ using the extended
%matrix formalism.  Let the length of ${\bf t}_*$ be $M$; then, the
%matrix $k({\bf t}_*,{\bf t})$ has a size $M \times N$, and so a total
%of $MN$ multiplications is required to obtain the predicted values.
%If $M$ is large, this can become quickly prohibitive.  It turns out
%that the structure of the sum of exponential kernels may be exploited
%to obtain the predicted mean in of order $p(M+N)+p^2N$ multiplications,
%i.e.\ a linear scaling with the total number of data points, which can
%be much fewer if $M$ is large.
%
%The procedure works as follows:
%\begin{itemize}
%\item Sort ${\bf t}$ and ${\bf t}_*$ in time order.
%\item Starting with the smallest value of ${\bf t}_*$, $t_{*,1}$,
%compute:
%\begin{eqnarray} \label{prediction}
%\bar f_{*,k} &=& \sum_p \alpha_p \left[\epsilon^+_{p,k} + \epsilon^-_{p,k}\right]\\
%\epsilon^+_{p,k} &=& \sum_{j \forall t_{*,k} > t_j} e^{-\beta_p (t_{*,k}-t_j)} h_j\\
%\epsilon^-_{p,k} &=& \sum_{j \forall t_{*,k} \le t_j} e^{-\beta_p (t_j-t_{*,k})} h_j,
%\end{eqnarray}
%where we have separated out the training times before and after the predicted time $t_{*,k}$.
%
%\item Next, we update the coefficients recursively:
%\begin{eqnarray}
%\epsilon^+_{p,k} &=& e^{-\beta_p (t_{*,k}-t_{*,k-1})} \left[\epsilon^+_{p,k-1} + \sum_{j \forall t_{*,k} > t_j \ge t_{*,k-1}} e^{-\beta_p (t_{*,k}-t_j)} h_j\right]\\
%\epsilon^-_{p,k} &=& e^{-\beta_p (t_{*,k}-t_{*,k-1})} \left[\epsilon^-_{p,k-1} - \sum_{j \forall t_{*,k} > t_j \ge t_{*,k-1}} e^{-\beta_p (t_j-t_{*,k})} h_j\right].
%\end{eqnarray}
%With these updates, we can then use equation \ref{prediction} to compute the expected
%mean at $t_{*,k}$ with an additional $p$ operations.  Since there are $N$ values of
%$t_k$, the number of operations needed to update $\epsilon$ is of order $p(N+M)$, and
%so this procedure avoids a heavy computational burden.
%
%The variance may be handled in a similar manner, but with $p^2$ coefficients
%to recursively update. For example, if $t_{*,k} > t_j \forall j,k$ then:
%\begin{eqnarray}
%cov(f_{*,k}) &=&  k(t_{*,k},t_{*,k}) - \sum_p \sum_q \alpha_p \alpha_q e^{-\beta_p(t_{*,k}-{\bf t}^T)} k({\bf t},{\bf t})^{-1} e^{-\beta_q (t_{*,k}-{\bf t})},\\
%&=& k(0) - \sum_p \sum_q \alpha_p \alpha_q \zeta_{p,q,k},\\
%\zeta_{p,q,k} &=&   e^{-\beta_p(t_{*,k}-{\bf t}^T)} k({\bf t},{\bf t})^{-1} e^{-\beta_q (t_{*,k}-{\bf t})},\\
%&=&  e^{-(\beta_p+\beta_q)(t_{*,k}-t_{*,k-1})} \zeta_{p,q,k-1}.
%\end{eqnarray}
%The recursive updates to the $\zeta_{p,q,k}$ parameters can be made at each step.
%\end{itemize}
%
%Problem:  if we have the case that the datapoints are interspersed, and if the sign changes,
%then we need to resolve $K^{-1} e^{-\beta_q \vert t_{*,k}-{\bf t}\vert}$.  However, maybe we can
%just solve for the difference in fewer operations? {\bf TBD}

@@

Rather than using $c_j+id_j$ and its complex conjugate as components of the kernel, we instead
find the real and imaginary components of the complex equation by taking the
sum and difference of the $c_j+id_j$ and $c_j-id_j$ equations (divided by 2 and $2i$, respectively),
and then discard the $c_j-id_j$ equation, which is redundant.  This keeps the same number of
additional equations for each $(c_j,d_j)$ with non-zero imaginary component (four additional
equations in the extended matrix), but converts the complex equations to real, thus avoiding
complex arithmetic, while the kernel components with $d_j=0$ still only have two
additional components in the extended matrix.  The extended matrix adds an extra band above
and below the diagonal when the $c_j+id_j$ and $c_j-id_j$ equations are combined since this
mixes the components of the original equations which are shifted by one column from one
another.  We will enumerate the number of $d_j=0$ cases as $p_0$, while the total number
of $c_j$'s is still $p$.  This gives a total number of equations to be solved with the
extended matrix of $N_{ex}= (4(p-p_0)+2p_0+1)\times(N-1)+1$, where $N$ is the number of data
points.

>>>>

It turns out that this kernel function can be made even more general by
introducing complex parameters $a_j \to a_j+i\,b_j$ and $c_j \to c_j+i\,d_j$.
In this case, the covariance function becomes
\begin{eqnarray}\eqlabel{genrp-kernel-complex}
k_\bvec{\alpha}(\tau_{nm}) = \sigma_n^2\,\delta_{nm} +
    \sum_{j=1}^J &&\left[
    \frac{1}{2}(a_j + i\,b_j)\,\exp\left(-(c_j+i\,d_j)\,\tau_{nm}\right)
        \right. \nonumber\\
    &&+\left.
    \frac{1}{2}(a_j - i\,b_j)\,\exp\left(-(c_j-i\,d_j)\,\tau_{nm}\right)
\right]
\end{eqnarray}
and, for this function, \eq{gp-likelihood} can still be solved with
$\mathcal{O}(J^2\,N)$ operations.
The details of this method and a few implementation considerations are
discussed in \app{implementation}.

\eq{genrp-kernel-complex} can also be expressed as
\begin{eqnarray}\eqlabel{genrp-kernel}
k_\bvec{\alpha}(\tau_{nm}) = \sigma_n^2\,\delta_{nm} +
    \sum_{j=1}^J &&\left[
    a_j\,\exp\left(-c_j\,\tau_{nm}\right)\,\cos\left(d_j\,\tau_{nm}\right)
        \right.\nonumber\\
    &&+ \left.
    b_j\,\exp\left(-c_j\,\tau_{nm}\right)\,\sin\left(d_j\,\tau_{nm}\right)
\right] \quad.
\end{eqnarray}
The Fourier transform of this covariance function is the power spectral
density (PSD) of the model and it is given by
\begin{eqnarray}\eqlabel{genrp-psd}
S(\omega) = \sum_{j=1}^J \sqrt{\frac{2}{\pi}}
\frac{(a_j\,c_j+b_j\,d_j)\,({c_j}^2+{d_j}^2)+(a_j\,c_j-b_j\,d_j)\,\omega^2}
{\omega^4+2\,({c_j}^2-{d_j}^2)\,\omega^2+({c_j}^2+{d_j}^2)^2}\quad.
\end{eqnarray}
The physical interpretation of this model isn't immediately obvious and we
will return to a more general discussion of the physical intuition in a moment
but we can start with a discussion of some useful special cases of this model.

If we set the imaginary amplitude $b_j$ for some component $j$ to zero, that
term of \eq{genrp-kernel} becomes
\begin{eqnarray}
k_j(\tau_{nm}) =
    a_j\,\exp\left(-c_j\,\tau_{nm}\right)\,\cos\left(d_j\,\tau_{nm}\right)
\end{eqnarray}
and the PSD for the this component is
\begin{eqnarray}
S_j(\omega) = \frac{1}{\sqrt{2\,\pi}}\,\frac{a_j}{c_j}\,\left[
    \frac{1}{1+\left(\frac{\omega-d_j}{c_j}\right)^2} +
    \frac{1}{1+\left(\frac{\omega+d_j}{c_j}\right)^2}
\right] \quad.
\end{eqnarray}
This PSD is the sum of two Lorentzian or Cauchy distributions with width
$c_j$ centered on $\omega = \pm d_j$.
This model can be interpreted intuitively as a quasiperiodic oscillator with
amplitude $A_j = a_j$, quality factor $Q_j = c_j$ \agoltodo{quality factor should
be dimensionless}, and period $P_j = 2\,\pi /
d_j$.

@@

In the case of $J$ damped sinusoids we can check for negative values of the PSD by solving for the
roots of the power spectrum, abbreviating with $z = \omega^2$:
\begin{equation}
P(\omega)=  \sum_{j=1}^J \frac{q_j z + r_j}{z^2+s_jz + t_j} = 0
\end{equation}
where
\begin{eqnarray}
q_j &=& a_jc_j-b_jd_j\\
r_j &=& (d_j^2+c_j^2)(b_jd_j+a_jc_j)\\
s_j &=& 2(c_j^2-d_j^2)\\
t_j &=& (c_j^2+d_j^2)^2.
\end{eqnarray}
The denominators of each term are positive, so we can multiply through by $\Pi_j \left(z^2+s_jz + t_j\right)$ yielding:
\begin{equation}
P_0(z) = \sum_{j=1}^J (q_j z + r_j)\Pi_{k \ne j}\left(z^2+s_kz + t_k\right) = 0,
\end{equation}
which is a polynomial with order $p_{ord}=2(J-1)+1$.  With $J=2$, this yields a cubic equation
which may be solved exactly for the roots.

@@

{\item Substituting $Q = 1/2$, \eq{sho-psd} becomes
\begin{eqnarray}
S(\omega) =
\sqrt{\frac{2}{\pi}}\,\frac{S_0}{\left[(\omega/\omega_0)^2+1\right]^2}
\end{eqnarray}
with the corresponding covariance function
\begin{eqnarray}
k(\tau) &=& \lim_{f \to 0}\,
    \frac{1}{2}\,S_0\,\omega_0\,e^{-\omega_0\,\tau}\,
    \left[\cos(f\,\tau) + \frac{\omega_0}{f}\,\sin(f\,\tau)\right] \\
&=& \frac{1}{2}\,S_0\,\omega_0\,e^{-\omega_0\,\tau}\,[1+\omega_0\,\tau] \quad.
\end{eqnarray}
This covariance function is also known as the Mat\'ern-3/2 function
\citep{Rasmussen:2006}.
}

@@

%The kernel is modeled as a sum of damped harmonics,
%\begin{equation}
%k_j(\tau) = a_0 \delta(\tau) + \sum_{j=1}^J a_j \exp{(-b_j \tau)} \cos{(-d_j \tau)} + b_j \exp{(-b_j \tau)} \sin{(-d_j \tau)},
%\end{equation}
%where $\tau > 0$ is a time-offset between data points, $c_j$ represents the decay
%in correlation with time and $d_j$ represents oscillations of period $P = 2\pi/d_j$, and
%$a_0 = \sigma^2$ is a white-noise component with $\delta(0) = 1$ and zero otherwise.  Formally
%the $\delta(\tau)$ function is a Kronecker delta function for which $\tau$ is the midpoint in
%time measured over a finite duration exposure time, and $\sigma$ is the standard deviation
%of the measured times if the correlated components are zero.

@@

One advantage of the Mat\'ern 3/2 kernel is that it is smooth across
$\tau=0$ since its derivative is zero at $\tau=0$.  The exponential kernel {\it does not}
have this property, which leads to qualitatively different noise properties on small
timescales due to the sharpness of the exponential kernel at $\tau=0$.  The Mat\'ern 3/2 kernel
turns out to be similar to the observed properties of stellar activity and granulation, and thus the approximate
formula is simultaneously physically relevant and computationally convenient.
The power spectrum of a single exponential scales as $f^{-2}$ at high frequency,
while the Mat\'ern kernel scales as $f^{-4}$ which more closely matches the power
spectrum due to activity and granulation in the Sun \citep{2004A&A...414.1139A,Michel:2009}.
\item {\bf Mat\'ern 5/2 kernel.}  The Mat\'ern 5/2 kernel is given by
\begin{equation}
k_{5/2}(t) = C_{5/2}(y) = \sigma^2 (1+y+y^2/3)e^{-y},
\end{equation}
where $y = \sqrt{5}\vert \tau/\tau_0\vert$.  The Fourier transform is:
\begin{equation}
\hat k_{5/2}(\omega) = \frac{2\tau_0 10^{5/2}}{3\pi^{1/2}} (5+\tau_0^2\omega^2)^{-3}.
\end{equation}
This kernel may be approximated well by the sum of four exponential kernels
\begin{eqnarray}
k_{5/2}(\tau) &\approx& -147.6965573 e^{-1.0097610 y} + 85.8875034 e^{-1.0621734 y}\\
 &-&4.6540017 e^{-0.8353745 y} + 67.4630744 e^{-0.9160394 y},
\end{eqnarray}
which is accurate to $<4\times 10^{-6}$ for $y \le 10$.
The Fourier transform has a steeper dependence at large frequencies, $f^{-6}$, and as we are not aware of a stellar
power spectrum that requires a steeper spectrum than this, we forego approximating further Mat\'ern kernels.
\item {\bf Exponential-sine-square kernel.}  This kernel is a periodic, positive valued kernel
which has been used to model quasi-periodic stellar variability due to star spots.   It is
given by $E(w,\Gamma) = e^{-\Gamma \sin^2{w}}$ where $w = \frac{\pi}{P}\vert \tau \vert$, which may be
expanded as a series in $\cos{(2kw)} = \cos{\frac{2\pi}{P} k \vert \tau \vert)}$,
with $k \in \mathbb{N}$.  We have expanded this to $k=4$, and fit for the coefficients, giving:
\begin{equation}
E(w,\Gamma) \approx \sum_{k=0}^4 \left(c_{k,1}+c_{k,2} e^{-c_{k,4}\Gamma^{c_{k,3}}}\right) \cos{(2kw)},
\end{equation}
with $c_{0,1}=0.24999, c_{0,2}=0.752803, c_{0,3}=0.960477, c_{0,4}=0.643393$ for $k=0$,
$c_{1,1}=0.450851, c_{1,2}=-0.445338, c_{1,3}=0.960477, c_{1,4}= 1.18102$ for $k=1$,
$c_{2,1}=0.206593, c_{2,2}=-0.207361, c_{2,3}=1.65902, c_{2,4}= 0.210641$ for $k=2$,
$c_{3,1}=0.0728449, c_{3,2}=-0.0729381, c_{3,3}=2.43149, c_{3,4}= 0.0471224$ for $k=3$, and
$c_{4,1}=0.019204, c_{4,2}=-0.0192129, c_{4,3}=3.27977, c_{4,4}= 0.0114198$ for $k=4$.
This approximation is accurate to better than 0.6\% for $\Gamma < 3$, with a standard
deviation of 0.1\%.
\end{itemize}

@@

%\subsection{Derivative of likelihood with respect to Kernel parameters}
%
%The derivative of the likelihood can be computed in $O(N)$ operations as follows.
%
%The derivative of the log likelihood with respect to model parameters ${\bf x}$ is given by:
%\begin{equation}
%    \frac{\partial \ln{\cal L}}{\partial x_i} = (\ln{\cal L})^\prime = -\frac{1}{2}\sum_i \frac{K_{ii}^\prime}{K_{ii}}
%    - \frac{1}{2} {\bf y}^T {\bf K}^{-1} {\bf K}^\prime {\bf K}^{-1} {\bf y},
%\end{equation}
%where $^\prime$ means the derivative wrt $x_i$.
%The kernel may be written as:
%\begin{equation}
%    K_{ij}({\bf \alpha},{\bf \beta}) = \left\{
%    \begin{array}{cc}
%       w_i + \sum_{l=1}^J \alpha_l & j = i \\
%       \sum_{l=1}^J \alpha_l \exp{(-\beta_l |t_i-t_j|)}  & j \ne i
%    \end{array}
%    \right.
%\end{equation}
%Since this equation is the sum of $J$ semi-separable matrices, then applying the product rule to compute the derivative will give a single semi-separable matrix for the derivative with respect to each kernel parameter.
%
%Now, letting ${\bf x} = (w,{\bf \alpha},{\bf \beta})$, then the derivative may be written as:
%\begin{eqnarray}
%    \frac{\partial K_{ii}}{\partial w} &=& \delta_{ij}\\
%    \frac{\partial K_{ij}}{\partial \alpha_k} &=&
%     \left\{
%    \begin{array}{cc}
%       1 & j = i \\
%        \exp{(-\beta_k |t_i-t_j|)}  & j \ne i
%    \end{array}
%    \right.\\
%    \frac{\partial K_{ij}}{\partial \beta_k} &=&
%     \left\{
%    \begin{array}{cc}
%       0  & j = i \\
%        -\alpha_k|t_i-t_j|\exp{(-\beta_k |t_i-t_j|)}  & j \ne i
%    \end{array}
%    \right.
%\end{eqnarray}
%The first term in the derivative of the likelihood can be computed in $O(N)$
%evaluations, while the second term can be computed in three steps:
%\begin{itemize}
%\item First, solve ${\bf z_1} = {\bf K}^{-1}{\bf y}$ with the GRP solver.
%\item Next, create an e
%\end{itemize}
%Writing this term out:

@@

%\section{Prediction/interpolation}
%
%Precition/interpolation involves constructing a matrix of covariances
%between the (noisy) training set and the times at which we wish to
%predict or interpolate data points.   If our training times are given by
%${\bf t}$, while the predicted times are given by ${\bf t}_*$, then
%mean values at the predicted times are given by (RW 2.23):
%\begin{eqnarray}
%\bar f_{*,k} &=& \sum_j k(t_{*,k},t_j) b_j,\\
%&=& \sum_j \sum_p \alpha_p e^{-\beta_p \vert t_{*,k}-t_j\vert} b_j,\\
%&=& \sum_p \alpha_p \sum_j e^{-\beta_p \vert t_{*,k}-t_j\vert} b_j,\\
%y_k &=& \sum_j k(t_j,t_k) h_j,
%\end{eqnarray}
%where the latter equation may be solved for ${\bf h}$ using the extended
%matrix formalism.  Let the length of ${\bf t}_*$ be $M$; then, the
%matrix $k({\bf t}_*,{\bf t})$ has a size $M \times N$, and so a total
%of $MN$ multiplications is required to obtain the predicted values.
%If $M$ is large, this can become quickly prohibitive.  It turns out
%that the structure of the sum of exponential kernels may be exploited
%to obtain the predicted mean in of order $J(M+N)+J^2N$ multiplications,
%i.e.\ a linear scaling with the total number of data points, which can
%be much fewer if $M$ is large.
%
%The procedure works as follows:
%\begin{itemize}
%\item Sort ${\bf t}$ and ${\bf t}_*$ in time order.
%\item Starting with the smallest value of ${\bf t}_*$, $t_{*,1}$,
%compute:
%\begin{eqnarray} \label{prediction}
%\bar f_{*,k} &=& \sum_p \alpha_p \left[\epsilon^+_{p,k} + \epsilon^-_{p,k}\right]\\
%\epsilon^+_{p,k} &=& \sum_{j \forall t_{*,k} > t_j} e^{-\beta_p (t_{*,k}-t_j)} h_j\\
%\epsilon^-_{p,k} &=& \sum_{j \forall t_{*,k} \le t_j} e^{-\beta_p (t_j-t_{*,k})} h_j,
%\end{eqnarray}
%where we have separated out the training times before and after the predicted time $t_{*,k}$.
%
%\item Next, we update the coefficients recursively:
%\begin{eqnarray}
%\epsilon^+_{p,k} &=& e^{-\beta_p (t_{*,k}-t_{*,k-1})} \left[\epsilon^+_{p,k-1} + \sum_{j \forall t_{*,k} > t_j \ge t_{*,k-1}} e^{-\beta_p (t_{*,k}-t_j)} h_j\right]\\
%\epsilon^-_{p,k} &=& e^{-\beta_p (t_{*,k}-t_{*,k-1})} \left[\epsilon^-_{p,k-1} - \sum_{j \forall t_{*,k} > t_j \ge t_{*,k-1}} e^{-\beta_p (t_j-t_{*,k})} h_j\right].
%\end{eqnarray}
%With these updates, we can then use equation \ref{prediction} to compute the expected
%mean at $t_{*,k}$ with an additional $J$ operations.  Since there are $N$ values of
%$t_k$, the number of operations needed to update $\epsilon$ is of order $J(N+M)$, and
%so this procedure avoids a heavy computational burden.
%
%The variance may be handled in a similar manner, but with $J^2$ coefficients
%to recursively update. For example, if $t_{*,k} > t_j \forall j,k$ then:
%\begin{eqnarray}
%cov(f_{*,k}) &=&  k(t_{*,k},t_{*,k}) - \sum_p \sum_q \alpha_p \alpha_q e^{-\beta_p(t_{*,k}-{\bf t}^T)} k({\bf t},{\bf t})^{-1} e^{-\beta_q (t_{*,k}-{\bf t})},\\
%&=& k(0) - \sum_p \sum_q \alpha_p \alpha_q \zeta_{p,q,k},\\
%\zeta_{p,q,k} &=&   e^{-\beta_p(t_{*,k}-{\bf t}^T)} k({\bf t},{\bf t})^{-1} e^{-\beta_q (t_{*,k}-{\bf t})},\\
%&=&  e^{-(\beta_p+\beta_q)(t_{*,k}-t_{*,k-1})} \zeta_{p,q,k-1}.
%\end{eqnarray}
%The recursive updates to the $\zeta_{p,q,k}$ parameters can be made at each step.
%\end{itemize}
%
%Problem:  if we have the case that the datapoints are interspersed, and if the sign changes,
%then we need to resolve $K^{-1} e^{-\beta_q \vert t_{*,k}-{\bf t}\vert}$.  However, maybe we can
%just solve for the difference in fewer operations? {\bf TBD}

@@

Rather than using $c_j+id_j$ and its complex conjugate as components of the kernel, we instead
find the real and imaginary components of the complex equation by taking the
sum and difference of the $c_j+id_j$ and $c_j-id_j$ equations (divided by 2 and $2i$, respectively),
and then discard the $c_j-id_j$ equation, which is redundant.  This keeps the same number of
additional equations for each $(c_j,d_j)$ with non-zero imaginary component (four additional
equations in the extended matrix), but converts the complex equations to real, thus avoiding
complex arithmetic, while the kernel components with $d_j=0$ still only have two
additional components in the extended matrix.  The extended matrix adds an extra band above
and below the diagonal when the $c_j+id_j$ and $c_j-id_j$ equations are combined since this
mixes the components of the original equations which are shifted by one column from one
another.  We will enumerate the number of $d_j=0$ cases as $J_0$, while the total number
of $c_j$'s is still $J$.  This gives a total number of equations to be solved with the
extended matrix of $N_{ex}= (4(J-J_0)+2J_0+1)\times(N-1)+1$, where $N$ is the number of data
points.
